String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-10-19 2023-1-19/sections/algorithms.tex
some updates

\medskip
\noindent
\emph{Step 1.} For each $i \in [n]$, compute a CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$, and let $\Lambda'_{x_i} := \{\cB_i\}$. \qed

@@

Since
$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}$ for each $i \in [n]$ and $\eta \in \parikh(\cU_{\cB_i})$, it turns out that 
the $\myvec{0}$-labeled transitions in $\cU_{\cB_i}$ do not contribute to the final output $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$. Therefore, we propose the following size-reduction technique for the NFA $\cU_{\cB_i}$'s.

\medskip
\noindent
\emph{Step 2.} For each $i \in [n]$, we view the transitions $(q, \myvec{0}, q')$ in $\cU_{\cB_i}$ as $\varepsilon$-transitions $(q, \varepsilon, q')$, and remove the $\varepsilon$-transitions from $\cU_{\cB_i}$. Then we determinize and minimize the resulting NFA.   \qed

@@

\medskip
\noindent
\emph{Step 3.} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then we solve the satisfiability of the following formula,
%
\vspace{-1mm}
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
\vspace{-1mm}
\qed
%\[
%\bigwedge \limits_{i \in [n]} \psi_{\cC_i}(\fZ_{\costset_{\cB_i}}) \wedge 
%\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
%\]

>>>>

\medskip
\noindent
\emph{Step 1.\denghang{(Intersection)}} For each $i \in [n]$, compute a CEFA $\cB_{i} = \aut_{i,1} \cap \cdots \cap \aut_{i, l_i}$, and let $\Lambda'_{x_i} := \{\cB_i\}$. \qed

@@

Since
$\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v} = \sum \limits_{\myvec{v} \in \costset(\cB_i) \setminus \{\myvec{0}\}} \eta(\myvec{v}) \myvec{v}$ for each $i \in [n]$ and $\eta \in \parikh(\cU_{\cB_i})$, it turns out that 
the $\myvec{0}$-labeled transitions in $\cU_{\cB_i}$ do not contribute to the final output $\sum \limits_{\myvec{v} \in \costset(\cB_i)} \eta(\myvec{v}) \myvec{v}$. \sout{Therefore, we propose the following size-reduction technique for the NFA $\cU_{\cB_i}$'s.} \denghang{Regardless of alphabet, we take register updates as new alphabet symbols and $\myvec{0}$ as $\epsilon$ symbol. Then we are capable of using standard minimization to reduce the automata size.}

\medskip
\noindent
\emph{Step 2.\denghang{(Automata size reduction)}} For each $i \in [n]$, we view the transitions $(q, \myvec{0}, q')$ in $\cU_{\cB_i}$ as $\varepsilon$-transitions $(q, \varepsilon, q')$, and remove the $\varepsilon$-transitions from $\cU_{\cB_i}$. Then we determinize and minimize the resulting NFA.   \qed

@@

\medskip
\noindent
\emph{Step 3.\denghang{(Parikh image)}} For each $i \in [n]$, we compute an existential LIA formula $\psi_{\cC_i}(\fZ_{\costset_{\cC_i}})$ from $\cC_i$ to represent $\parikh(\cC_i)$. Then we solve the satisfiability of the following formula,
%
\vspace{-1mm}
\[
\begin{array}{l}
\bigwedge \limits_{i \in [n]} \left(\psi_{\cC_i}(\fZ_{\costset_{\cC_i}}) \wedge \alpha_i\left[\sum\limits_{\myvec{v} \in \costset_{\cC_i}} \anivarz_{i, \myvec{v}} \myvec{v} \big{/}R_{\cB_i}  \right] \right) \ \wedge \\
\ \ \varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cC_1}} \anivarz_{1, \myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cC_n}} \anivarz_{n, \myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
\end{array}
\]
\vspace{-1mm}
\qed
%\[
%\bigwedge \limits_{i \in [n]} \psi_{\cC_i}(\fZ_{\costset_{\cB_i}}) \wedge 
%\varphi \left[ \sum\limits_{\myvec{v} \in \costset_{\cB_1}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_1}, \cdots, \sum\limits_{\myvec{v} \in \costset_{\cB_n}} \anivarz_{\myvec{v}} \myvec{v} \big{/}R_{\cB_n} \right].
%\]

### 2023-10-19 2023-1-19/sections/cefa.tex
some updates

In the sequel, we define three CEFA operations: union, intersection, and concatenation. The following section will use these three operations to solve RECL constraints. \denghang{Note that the union, intersection, and concatenation operations are defined in a slightly more involved manner than symbolic register automata~\cite{sra} and counter automata~\cite{GGM12}, as a result of the technical choice that the accepting conditions of CEFA do not distinguish final states.}

>>>>

In the sequel, we define three CEFA operations: union, intersection, and concatenation. The following section will use these three operations to solve RECL constraints. \denghang{Note that the union, intersection, and concatenation operations are defined in a slightly more involved manner than register automata~\cite{ra} and counter automata~\cite{GGM12}, as a result of the technical choice that the accepting condition is attached to each automaton, not each final state.}

### 2023-10-19 2023-1-19/sections/implementation.tex
some updates

%!TEX root = ../main.tex
%\documentclass{standalone}
%\begin{document}
In this section, we list our experimental results. We first present the overall experiment results on RegCoL and AutomatArk benchmark suites. Then we present experiment results of the large-counting part and evaluate the simplification techniques. \denghang{The goal of our experiments is to answer the following questions:
\begin{itemize}
  \item [Q1:] Are our decision procedure more efficient than the state-of-the-art string constraint solvers on RECL constraints? (Sec \ref{subsec:overall_eval})
  \item [Q2:] Why is our decision procedure more efficient? (Sec \ref{subsec:large_bounds_eval})
  \item [Q3:] Are the size-reduction techniques practical? (Sec \ref{subsec:size_reduction_eval})
  \item [Q4:] How effective is the algorithm for solving the nonemptiness problem compared to the case that uses the nuXmv model checker~\cite{atva2020}? (Sec \ref{subsec:size_reduction_eval})
\end{itemize}
}

@@

We evaluate the performance of $\ostrichrecl$ on 1,969 instances of the Large-Bounds part. 
The experiment results are put in Fig~\ref{fig:table_large_count}. From the results, $\ostrichrecl$ solves \textbf{1,873} instances correctly, which is \textbf{947/278/563/637/523 more} than that solved by CVC5/Z3str3RE/Z3str3/Z3seq/OSTRICH respectively. Moreover, $\ostrichrecl$ is \textbf{ 6.79/2.88/2.61/5.27/3.95} times faster than CVC5/ Z3str3RE/Z3str3/Z3seq/OSTRICH respectively. \denghang{To answer \textbf{Q2}, we can see that $\ostrichrecl$ is far more efficient than the other solvers on problem instances with large bounds. \textbf{It is because the complexity in our decision procedure is irrelevant to counting bounds}, while the complexity of other solvers is linear or even exponential to counting bounds.}

\subsection{Evaluation of the effectiveness of size-reduction techniques}\label{subsec:size_reduction_eval}
% \medskip
% \noindent 
% \emph{Evaluation of the effectiveness of the automata size-reduction techniques.}
We also do experiments to evaluate the effectiveness of the automata size-reduction techniques, that is, Step 2 in Section~\ref{subsec:cefadec}. 
Let us use $\ostrichrecl_{\rm -SIMP}$ to denote $\ostrichrecl$ with the size-reduction techniques removed. 
Meanwhile, we compare $\ostrichrecl$ and $\ostrichrecl_{\rm -SIMP}$ against $\ostrichrecl_{\rm NUXMV}$, where the $\cefadec$ problem is solved by the \textsc{nuXmv} model checker \cite{nuxmv}, instead of the procedure in Section~\ref{subsec:cefadec}. 
The experiment results are put in Table~\ref{tab:results_simp}. 
%$\ostrichrecl_{\rm -SIMP}$ is the tool without simplification techniques. $\ostrichrecl_{\rm NUXMV}$ is the tool using \textsc{nuXmv}-based techniques. 
From the results, $\ostrichrecl$ solves \textbf{1,503 more} instances and is \textbf{2.21} times faster than $\ostrichrecl_{\rm -SIMP}$. \denghang{To answer \textbf{Q3}, \textbf{the size-reduction techniques indeed play a vital role in the performance improvement}}. Moreover, $\ostrichrecl$ solves \textbf{1,798 more} instances and is \textbf{3.13} times faster than $\ostrichrecl_{\rm NUXMV}$. \denghang{To answer \textbf{Q4}, \textbf{our algorithm for solving the nonemptiness problem is much better than the \textsc{nuXmv}-based techniques}}.

>>>>

%!TEX root = ../main.tex
%\documentclass{standalone}
%\begin{document}
In this section, we list our experimental results. We first present the overall experiment results on RegCoL and AutomatArk benchmark suites. Then we present experiment results of the large-counting part and evaluate the simplification techniques. \denghang{The goal of our experiments is to answer the following questions:
\begin{itemize}
  \item [Q1:] Does the decision procedure perform well~(Sec \ref{subsec:overall_eval}), especially when the counting bounds are large?~(Sec \ref{subsec:large_bounds_eval})
  \item [Q2:] Are the size-reduction techniques practical?~(Sec \ref{subsec:size_reduction_eval})
  \item [Q3:] How effective is the algorithm for solving the nonemptiness problem compared to the case that uses the nuXmv model checker~\cite{atva2020}?~(Sec \ref{subsec:size_reduction_eval})
\end{itemize}
}

@@

We evaluate the performance of $\ostrichrecl$ on 1,969 instances of the Large-Bounds part. 
The experiment results are put in Fig~\ref{fig:table_large_count}. From the results, $\ostrichrecl$ solves \textbf{1,873} instances correctly, which is \textbf{947/278/563/637/523 more} than that solved by CVC5/Z3str3RE/Z3str3/Z3seq/OSTRICH respectively. Moreover, $\ostrichrecl$ is \textbf{ 6.79/2.88/2.61/5.27/3.95} times faster than CVC5/ Z3str3RE/Z3str3/Z3seq/OSTRICH respectively. \denghang{To answer \textbf{Q1}, we can see that $\ostrichrecl$ \textbf{is far more efficient than the other solvers on problem instances with large bounds.}}

\subsection{Evaluation of the effectiveness of size-reduction techniques}\label{subsec:size_reduction_eval}
% \medskip
% \noindent 
% \emph{Evaluation of the effectiveness of the automata size-reduction techniques.}
We also do experiments to evaluate the effectiveness of the automata size-reduction techniques, that is, Step 2 in Section~\ref{subsec:cefadec}. 
Let us use $\ostrichrecl_{\rm -SIMP}$ to denote $\ostrichrecl$ with the size-reduction techniques removed. 
Meanwhile, we compare $\ostrichrecl$ and $\ostrichrecl_{\rm -SIMP}$ against $\ostrichrecl_{\rm NUXMV}$, where the $\cefadec$ problem is solved by the \textsc{nuXmv} model checker \cite{nuxmv}, instead of the procedure in Section~\ref{subsec:cefadec}. 
The experiment results are put in Table~\ref{tab:results_simp}. 
%$\ostrichrecl_{\rm -SIMP}$ is the tool without simplification techniques. $\ostrichrecl_{\rm NUXMV}$ is the tool using \textsc{nuXmv}-based techniques. 
From the results, $\ostrichrecl$ solves \textbf{1,503 more} instances and is \textbf{2.21} times faster than $\ostrichrecl_{\rm -SIMP}$. \denghang{To answer \textbf{Q2}, \textbf{the size-reduction techniques indeed play a vital role in the performance improvement}}. Moreover, $\ostrichrecl$ solves \textbf{1,798 more} instances and is \textbf{3.13} times faster than $\ostrichrecl_{\rm NUXMV}$. \denghang{To answer \textbf{Q3}, \textbf{our algorithm for solving the nonemptiness problem is much better than the \textsc{nuXmv}-based techniques}}.

### 2023-10-19 2023-1-19/sections/introduction.tex
some updates

\medskip
\noindent
\emph{Related work.} 
We discuss more related work about regexes with counting in programming languages.  
The determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to quickly match a subclass of regexes with counting. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the nonemptiness problem of these automata was not considered, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length.
Real-world regexes in programming languages include features beyond classical regular expressions, e.g., the greedy/lazy Kleene star, capturing groups, and back references. \denghang{Symbolic register automata\cite{sra} are raised to handle back references.} Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving~\cite{CF+22}. Nevertheless, the counting operators are still unfolded explicitly in~\cite{CF+22} \denghang{and \cite{sra}}.  Solving string constraints involving lengths has also been investigated by various researchers, see e.g. \cite{WC+18,atva2020,z3str3re}. \denghang{Rule-based string solvers like Z3str/2/3/4 and CVC4/5 are incapable of handling regexes with counting and complementary, even for the simple constraint $x\in [0-8]^{\{1,200\}}\wedge x\not\in [0-9]^+$. Automata-based string solvers like Z3Str3RE and Ostrich can solve it, but they are inefficient because they have to build automata by unwinding counting operator, resulting in an automaton with more than $1800$ states for regex $[0-8]^{\{1,200\}}$. Moreover, the counting operator is not supported by string solvers Z3-Trau and Slent.}

>>>>

\medskip
\noindent
\emph{Related work.} 
We discuss more related work about regexes with counting in programming languages.  
The determinism of regexes with counting was investigated in~\cite{GGM12,CL15}. Counting-set automata were proposed in~\cite{redos_lenka,HS+23} to quickly match a subclass of regexes with counting. Moreover, a variant of nondeterministic counter automata, called bit vector automata, was proposed recently in \cite{GKM23} to enable fast matching of a more expressive class of regexes with counting.   Nevertheless, the nonemptiness problem of these automata was not considered, and it is unclear whether these automata models can be used for solving string constraints with regex-counting and string-length. \denghang{For the automata model, the definition of registers of CEFAs resembles the counters of the counter automata class~\cite{GGM12,redos_lenka,HS+23,GKM23}, but CEFAs contain no guards in each transition. Register automata~\cite{ra} and symbolic register automata~\cite{sra} utilize registers to store input variables (i.e., the letters of the alphabet), on which only equality and inequality are permitted. In contrast, CEFAs utilize registers to store integers constrained by linear arithmetic formulae.}
Real-world regexes in programming languages include features beyond classical regular expressions, e.g., the greedy/lazy Kleene star, capturing groups, and back references. Real-world regexes have been addressed in symbolic execution of Javascript programs~\cite{LMK19} and string constraint solving~\cite{CF+22}. Nevertheless, the counting operators are still unfolded explicitly in~\cite{CF+22}. \sout{Solving string constraints involving lengths has also been investigated by various researchers, see e.g. \cite{WC+18,atva2020,z3str3re}.} \denghang{Except for the string solvers mentioned before, some related works about string constraints with the length function exist. Trau~\cite{z3trau} relies on a Counter-Example Guided Abstraction Refinement (CEGAR) framework in which the under- and over-approximation of the automata are iteratively refined to solve length and regex constraints. Slent~\cite{WC+18} transfers the string and length constraints to length-encoded automata, and leverages a symbolic model checker as an engine for the emptiness checking of automata. However, both Trau and Slent do not support counting operators.}
